---
title: "Final Project Proposal for DATA 608"
author: "Thomas Hill"
output:
  html_document: default
  html_notebook: default
---




Final Project Proposal: New York State Maternity and Labor Statistics



Maternity and labor are a crucial time for many health interventions, and New York state notably mandates basic statistics be available for a few dozen characteristics of delivery. The route of delivery, fetal presentation, and use of anesthesia are all critical elements of the process, and understanding the prevalence and trends of each method may offer improvements in expectations to practitioners as well as laypeople.


```{r, csv}

ny_maternity <- read.csv('https://raw.githubusercontent.com/hillt5/DATA608/main/Hospital_Maternity_Information__Beginning_2008.csv')


```


The dataset available for NY state offers the required statistics for a period 2008 – 2017, separated by each hospital and its location. For this project, my goals are to provide some visualizations and apps for laypeople to explore maternity statistics.


```{r, colnames}

head(ny_maternity)

colnames(ny_maternity)
```


```{r, unique-entries}

unique(ny_maternity$Year)

length(unique(ny_maternity$Facility.ID))

unique(ny_maternity$Measure.Name)

unique(ny_maternity$Denominator)

unique(ny_maternity$Category)


```

Looking at the structure of the dataframe, it appears to be in a tall format. The dataframe has entries for 2010 - 2017, and it looks at 29 different measures collectted from 145 facilities. There are also 7 different categories consiered, while the Denominator category that describes where the percentages were calculated.


Previous projects have concentrated geographical visualizations of births, which mirrored overall population. In addition to a map, I also intend on adding webapps for visualizing trends over time, with adjustable time period and locations. I think it will also be useful to allow comparisons of nearby hospitals to see distinctions over smaller areas. Lastly, I’ll provide static graph offering comparisons of county population versus fertility and birth defect rate.



Previous projects: 

https://nycdatascience.com/blog/student-works/visualizing-birth-and-maternity-trends-in-new-york-state/


Source dataset:

https://health.data.ny.gov/Health/Hospital-Maternity-Information-Beginning-2008/net3-iygw



Potential adjunct datasets: 

https://health.data.ny.gov/Health/Birth-Defect-Prevalence-Beginning-1992/mz8x-255x
 

https://health.data.ny.gov/Health/Vital-Statistics-Live-Births-and-Fertility-Rates-b/c2tx-jecb
 

https://data.ny.gov/Government-Finance/Annual-Population-Estimates-for-New-York-State-and/krt9-ym2k

